A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-06-10 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
Fixed typos and grammar
using Grammarly

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context, and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math are needed. For the more philosophically inclined, we will point to areas where further conceptual work is needed.

@@

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.

We reformulate the problem in the following way: given the space of all possible statistical ensembles allowed by a statistical theory, what measure theoretic tools can we use to represent each ensemble and the geometric properties of the overall space? Why exactly classical probability measures fail to capture completely quantum states? What suitable generalizations are possible, and what mathematical tools would need to be developed?

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review some results from Choquet theory, which is, to our knowledge, the most general approach to represent elements of a convex set using additive probability measures. The key insight is that, at least in the compact case, such representation is always possible, but it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex and is a particularly shaped subset of a vector space. The space of quantum ensembles is not a simplex for one key reason: it allows multiple decompositions in terms of pure states. These insights will allow to better understand exactly the limits of classical probability to describe ensembles in physical theories.

We will also review some results from the use of quasi-probability measures. While the Wigner function represents the most known instance, there are different choices. Note that the space of signed probability measures over a set $K$ is a vector space, so one can find an affine map between a suitable space of signed probability measures and the embedding vector space. Each ensemble, then, is uniquely represented with a signed probability measure. However, unlike in the Choquet case, the measure is not supported by the set of extreme points of the ensemble space. That is, there is no unique or privileged choice for this representation. Moreover, since the map is at the level of the full vector space, not just the subset with valid ensembles, there is no general way to characterize which subset of signed probability measures correspond to ensembles. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always provide a physically meaningful representation. Moreover, in the compact case, these measures can be restricted to the extreme points while retaining uniqueness. (TODO: prove)

The fraction capacity is the supremum of all possible measures for the same ensemble. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure, and is therefore additive. The state capacity, instead, corresponds to the minimum of the entropy of all possible representations. It recovers the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Having shown that a general approach based on non-additive measures is possible, we lay down a series of questions to explore in the future. The biggest one is whether it is possible to generalize notions such as integration, derivation and expectation to this non-additive case.  It should be noted that non-additive measures are used in the context of decision theory, so there may be a connection between the use of non-additive measures in quantum theory and interdependent complimentary choices.[TODO: find citation that sub-additive is when choices are complementary] Both the Choquet and the Sugeno integral, for example, are defined over a monotonic set-functions, so there is a chance that these tools may turn out to be useful. We see this work as starting this line of research, which will take more effort to investigate.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects. Therefore, it is important to us that this work is a piece of a much larger puzzle, which connects cleanly and directly to  other areas, such as information geometry, functional analysis, quantum logic and so on. It is the hallmark of a true fundamental structure for physics that such connections can be easily made. Unfortunately, the nature of scientific publication requires us to publish a narrow slice of the work to fit the scope of a particular journal. Ironically, this means that the most important motivations for our approach are out of scope of this article, as we will concentrate only to the connections to measure theory.

@@

[This section has to make clear in what extra conditions on ensembles spaces Choquet theory can be applied.]

@@

% Physical meaning of representation with a measure.
First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

% Limitation of applying Choquet theory to our case
We should now look at the requirements of the theorem, to see whether we can always apply it to an ensemble space. First of all, and ensemble space $\Ens$ is a convex subset of a convex sapce $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measure over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore the theorem does not apply in general.

% Define "Choquet space" as those spaces where we can apply Choquet theory
We can apply Choquet's theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allow a representation through a probability on points that may correspond to idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

Note that ensembles over classical spaces will be represented by a unique probability measure, therefore the corresponding closure of the ensemble space is a Choquet simplex. However, quantum mixed states can be decomposed in multiple ways in terms of pure states, and therefore the closure of a quantum ensemble space is not a simplex.

@@

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a topological space $X$, a measure $\mu: \mathcal{B}(X) \to \mathbb{R}$ is affine if $\mu(X)=1$. All probability measures are affine measures, but the reverse is not true as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

Given a space $X$, the space of all affine measure is a vector space, not just a convex subset as in the case of the space of probability measures. This means that, given an ensemble space, we may find a space $X$ such that the space of affine measures over $X$ is isomorphic to the embedding vector space $V$ as topological vector spaces. In this case, any element $v\in V$ can be represented uniquely by an affine measure $\mu_{v}$, regardless of whether they are actual ensembles or not.\footnote{The question of whether such an $X$ always exists is left open. We suspect a link exists between the statistical quantities $F_i$ that fully identify each ensemble and the space $X$.} Because the statistical quantities $F_i$ are affine functions, we will have $F_i(v) = \int_X F_i d\mu_{v}$.

@@

Given that mixing coefficients have a more general applicability than probability of outcomes, let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probably for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definitions.

Given a target ensemble $\ens \in \Ens$ and an arbitrary ensemble $\ens[a] \in \Ens$, we define the \textbf{fraction} of $\ens[a]$ in $\ens$ to be
\begin{equation}
	\fraction_{\ens}(\ens[a]) = \sup(\{ p \in [0,1] \, | \, \exists \, \ens_1 \in \Ens \text{ s.t. }  \ens = p \ens[a] + \bar{p} \ens_1 \}).
\end{equation}
The fraction is always well-defined because we can always write $\ens = 0 \ens[a] + 1 \ens$, therefore it must be zero or greater. This quantity tells us how much of ensemble $\ens$ can be constructed from $\ens[a]$.

We now extend this idea from a single ensemble $\ens[a] \in \Ens$ to a Borel set of ensembles $A \subset \Ens$. Given that the $\hull(A)$ is the set of all possible convex combinations (i.e. mixtures),\footnote{Since we work with convex sets within a topology, the hull is both the closed under the topology and convex combinations. That is, $\hull(A)$ is the smallest closed and convex set that contains $A$} we define the \textbf{fraction capacity} of $A$ for $\ens$ to be
\begin{equation}
	\frcap_{\ens}(A) = \sup(\fraction_{\ens}(\hull(A))\cup\{0\}).
\end{equation}
This returns the biggest fraction of $\ens$ that can be achieved with a mixture of elements of $A$.\footnote{The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature. TODO \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} }

@@

The first property readily comes from the range of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures, therefore they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit. If not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure.

Our setup is very similar to what happens in Choquet theory. The fraction capacity, in fact, is not just defined on the extreme points, but it is defined on the whole space of ensembles. In the classical discrete case, for example, if $A \subset \Omega \subset \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed is a mixture of those distributions. Similarly, the probability measures in Choquet theory are, in general, defined on the whole ensemble space, and then one finds those particular ones that have support on the extreme points.

@@

As we saw, the classical case corresponds to a simplex in Choquet theory. This means that each element of the ensemble space is represented by a unique probability measure over the extreme points. In that case, the fraction capacity extended on the extreme points will also be additive. In the non-classical case, the closure of the ensemble space will not be a simplex, and there will be multiple representations, which will lead to a non-additive measure. Suppose, in fact, that $\mu$ and $\nu$ are two distinct measures that represent the same ensemble $\ens$. Then there will be a set $A$ such such that $\mu(A) \neq \nu(A)$. Suppose, without loss of generality, that $\mu(A) > \nu(A)$ We will also have $\mu(A^{\complement}) = 1- \mu(A) < 1 - \nu(A) = \nu(A^{\complement})$. Since the fraction capacity for a set cannot be less than the measure on that set, we have $\frcap_{\ens}(A \cup A^{\complement})=\frcap_{\ens}(X) = 1 = \mu(A) + \mu(A^{\complement}) < \mu(A) + \nu(A^{\complement}) \leq \frcap_{\ens}(A) + \frcap_{\ens}(A^{\complement})$. That is, the fraction capacity is additive over disjoint sets of extreme points in and only in the classical case. In other words, the fraction capacity reduces to a classical probability distribution over pure states in the classical case.

@@

We now want to introduce a measure theoretic representation of the entropy that generalizes the notion of count of states. Since entropy characterizes the variability of the preparations of the ensemble, greater variability means the ensemble is spread over more cases. Therefore the count of distinguishable states must be a monotonic function of the entropy. It has been shown[cite Hall] under very general conditions that the exponential of the entropy is the only function to have the appropriate features. One required feature, that we prove in general, is that the exponential of the entropy is subadditive (i.e. the count of states associated to a mixture cannot exceed the combined count of states associated to both mixtures) and it is additive for orthogonal ensembles (i.e. no overlap in possible instances) in a specific linear combination.

@@

The first property readily comes from the range of the exponential. For the second, note that $\hull$, exponential and $\sup$ are all monotone. For sub-additivity, the exponential of the entropy itself is sub-additive, and it is additive over orthogonal ensembles. For the last, since the entropy itself  is continuous, this property carries over to the set function.

Again we note that the non-additive measure we recover is very well-behaved, and connects directly with important quantities. When Choquet theory applies, state capacity can be extended to the extreme points. In the classical case, disjoint sets of extreme points will be orthogonal, meaning that the state capacity there restricted is an additive measure. In the discrete case, every point is associated to zero entropy, and since $2^0=1$, the state capacity reduces to the counting measure. In the continuous case, gives a subset of phase space $A$, the maximum entropy is reached by the uniform distribution $\rho_A$, which means we recover $\log_2 \capacity(A) = S(\rho_A)$, the fundamental postulate of statistical mechanics. Similarly, in quantum mechanics, given an orthogonal subspace $A$, the maximum entropy is recovered by the maximally mixed state, which means $\capacity(A) = 2^{\dim(A)}$.

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable. In quantum mechanics, only orthogonal state are distinguishable, and therefore the state capacity recovers the number of orthogonal states.

>>>>

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics, and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math is needed. For the more philosophically inclined, we will point to areas where further conceptual work is needed.

@@

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used, given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.

We reformulate the problem in the following way: given the space of all possible statistical ensembles allowed by a statistical theory, what measure theoretic tools can we use to represent each ensemble and the geometric properties of the overall space? Why exactly do classical probability measures fail to capture completely quantum states? What suitable generalizations are possible, and what mathematical tools would need to be developed?

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurement (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review some results from Choquet theory, which is, to our knowledge, the most general approach to represent elements of a convex set using additive probability measures. The key insight is that, at least in the compact case, such a representation is always possible, but it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex and is a particularly shaped subset of a vector space. The space of quantum ensembles is not a simplex for one key reason: it allows multiple decompositions in terms of pure states. These insights will allow to better understand exactly the limits of classical probability to describe ensembles in physical theories.

We will also review some results from the use of quasi-probability measures. While the Wigner function represents the most known instance, there are different choices. Note that the space of signed probability measures over a set $K$ is a vector space, so one can find an affine map between a suitable space of signed probability measures and the embedding vector space. Each ensemble, then, is uniquely represented with a signed probability measure. However, unlike in the Choquet case, the measure is not supported by the set of extreme points of the ensemble space. That is, there is no unique or privileged choice for this representation. Moreover, since the map is at the level of the full vector space, not just the subset with valid ensembles, there is no general way to characterize which subset of signed probability measures corresponds to ensembles. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of the count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always provide a physically meaningful representation. Moreover, in the compact case, these measures can be restricted to the extreme points while retaining uniqueness. (TODO: prove)

The fraction capacity is the supremum of all possible measures for the same ensemble. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure and is therefore additive. The state capacity, instead, corresponds to the minimum of the entropy of all possible representations. It recovers the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Having shown that a general approach based on non-additive measures is possible, we lay down a series of questions to explore in the future. The biggest one is whether it is possible to generalize notions such as integration, derivation, and expectation to this non-additive case.  It should be noted that non-additive measures are used in the context of decision theory, so there may be a connection between the use of non-additive measures in quantum theory and interdependent complementary choices.[TODO: find citation that sub-additive is when choices are complementary] Both the Choquet and the Sugeno integral, for example, are defined over a monotonic set-functions, so there is a chance that these tools may turn out to be useful. We see this work as starting this line of research, which will take more effort to investigate.

A final remark before we start. Too often, works in the foundations of physics focus on a particular aspect without concern for a broader connection to all relevant mathematical, physical, and philosophical aspects. Therefore, it is important to us that this work is a piece of a much larger puzzle, which connects cleanly and directly to other areas, such as information geometry, functional analysis, quantum logic, and so on. It is the hallmark of a true fundamental structure for physics that such connections can be easily made. Unfortunately, the nature of scientific publication requires us to publish a narrow slice of the work to fit the scope of a particular journal. Ironically, this means that the most important motivations for our approach are out of scope of this article, as we will concentrate only on the connections to measure theory.

@@

[This section has to make clear in what extra conditions on ensemble spaces Choquet theory can be applied.]

@@

% Physical meaning of representation with a measure.
First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows us to think of $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choquet simplex by the property of having a single decomposition in terms of extreme points.

% Limitation of applying Choquet theory to our case
We should now look at the requirements of the theorem to see whether we can always apply it to an ensemble space. First of all, an ensemble space $\Ens$ is a convex subset of a convex space $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measures over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore, the theorem does not apply in general.

% Define "Choquet space" as those spaces where we can apply Choquet theory
We can apply Choquet's theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allow a representation through a probability on points that may correspond to idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distributions over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

Note that ensembles over classical spaces will be represented by a unique probability measure, therefore the corresponding closure of the ensemble space is a Choquet simplex. However, quantum mixed states can be decomposed in multiple ways in terms of pure states, and therefore, the closure of a quantum ensemble space is not a simplex.

@@

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a topological space $X$, a measure $\mu: \mathcal{B}(X) \to \mathbb{R}$ is affine if $\mu(X)=1$, where $\mathcal{B}(X)$ is the Borel $\sigma$-algebra. All probability measures are affine measures, but the reverse is not true, as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

Given a space $X$, the space of all affine measures is a vector space, not just a convex subset as in the case of the space of probability measures. This means that, given an ensemble space, we may find a space $X$ such that the space of affine measures over $X$ is isomorphic to the embedding vector space $V$ as topological vector spaces. In this case, any element $v\in V$ can be represented uniquely by an affine measure $\mu_{v}$, regardless of whether they are actual ensembles or not.\footnote{The question of whether such an $X$ always exists is left open. We suspect a link exists between the statistical quantities $F_i$ that fully identify each ensemble and the space $X$.} Because the statistical quantities $F_i$ are affine functions, we will have $F_i(v) = \int_X F_i d\mu_{v}$.

@@

Given that mixing coefficients have a more general applicability than probability of outcomes, let us concentrate on the following problem: given an ensemble $\ens \in \Ens$ and a (Borel) set of ensembles $A \subseteq \Ens$, what fraction of $\ens$ can be constructed by a mixture of $A$? How much of $\ens$ can be explained as coming from preparations corresponding to $A$? For example, let $\Ens$ be the space of all probability distributions for a six-sided die. Let $\ens_{123456}$ be the uniform distribution over all outcomes. Let $A_1 = \{\ens_1\}$ where $\ens_1$ represents outcome one with 100\% probability. Since we can write $\ens_{123456} = \frac{1}{6} \ens_{1} + \frac{5}{6} \ens_{23456}$, where $\ens_{23456}$ represents a uniform distributions over the five outcomes, $1/6$ of $\ens$ can be constructed from $A$, but no more. Similarly, if $A_{12} = \{\ens_{1},\ens_{2}\}$, we can write $\ens = \frac{1}{3} \left(\frac{1}{2} \ens_1 + \frac{1}{2} \ens_2 \right)  + \frac{2}{3} \ens_{3456}$, so $1/3$ of $\ens$ can be constructed from $A$, but no more. Note that, given the uniform distribution, $1/6$ is exactly the probability for the event $A_1$ and $1/3$ the probability for event $A_{12}$. This gives the basic insight for our definitions.

Given a target ensemble $\ens \in \Ens$ and an arbitrary ensemble $\ens[a] \in \Ens$, we define the \textbf{fraction} of $\ens[a]$ in $\ens$ to be
\begin{equation}
	\fraction_{\ens}(\ens[a]) = \sup(\{ p \in [0,1] \, | \, \exists \, \ens_1 \in \Ens \text{ s.t. }  \ens = p \ens[a] + \bar{p} \ens_1 \}).
\end{equation}
The fraction is always well-defined because we can always write $\ens = 0 \ens[a] + 1 \ens$, therefore it must be zero or greater. This quantity tells us how much of the ensemble $\ens$ can be constructed from $\ens[a]$.

We now extend this idea from a single ensemble $\ens[a] \in \Ens$ to a Borel set of ensembles $A \subset \Ens$. Given that the $\hull(A)$ is the set of all possible convex combinations (i.e. mixtures),\footnote{Since we work with convex sets within a topology, the hull is both the closure under the topology and convex combinations. That is, $\hull(A)$ is the smallest closed and convex set that contains $A$} we define the \textbf{fraction capacity} of $A$ for $\ens$ to be
\begin{equation}
	\frcap_{\ens}(A) = \sup(\fraction_{\ens}(\hull(A))\cup\{0\}).
\end{equation}
This returns the biggest fraction of $\ens$ that can be achieved with a mixture of elements of $A$.\footnote{The name fraction capacity is chosen to both signify the ability of the set $A$ to contain $\ens$ (i.e. fraction capacity of one means $\ens$ is within the convex combinations of $A$) and to indicate that it will be a non-additive monotonic measure, which are called ``capacities'' in some literature. TODO \href{ https://link.springer.com/book/10.1007/978-3-319-03155-2}{cite} }

@@

The first property readily comes from the range of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures; therefore, they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit. If not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure.

Our setup is very similar to what happens in Choquet theory. The fraction capacity, in fact, is not just defined on the extreme points, but it is defined on the whole space of ensembles. In the classical discrete case, for example, if $A \in \Omega \ni \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $\frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed as a mixture of those distributions. Similarly, the probability measures in Choquet theory are, in general, defined on the whole ensemble space, and then one finds those particular ones that have support on the extreme points.

@@

As we saw, the classical case corresponds to a simplex in Choquet theory. This means that each element of the ensemble space is represented by a unique probability measure over the extreme points. In that case, the fraction capacity, extended to the extreme points, will also be additive. In the non-classical case, the closure of the ensemble space will not be a simplex, and there will be multiple representations, which will lead to a non-additive measure. Suppose, in fact, that $\mu$ and $\nu$ are two distinct measures that represent the same ensemble $\ens$. Then there will be a set $A$ such that $\mu(A) \neq \nu(A)$. Suppose, without loss of generality, that $\mu(A) > \nu(A)$ We will also have $\mu(A^{\complement}) = 1- \mu(A) < 1 - \nu(A) = \nu(A^{\complement})$. Since the fraction capacity for a set cannot be less than the measure on that set, we have $\frcap_{\ens}(A \cup A^{\complement})=\frcap_{\ens}(X) = 1 = \mu(A) + \mu(A^{\complement}) < \mu(A) + \nu(A^{\complement}) \leq \frcap_{\ens}(A) + \frcap_{\ens}(A^{\complement})$. That is, the fraction capacity is additive over disjoint sets of extreme points in and only in the classical case. In other words, the fraction capacity reduces to a classical probability distribution over pure states in the classical case.

@@

We now want to introduce a measure theoretic representation of the entropy that generalizes the notion of count of states. Since entropy characterizes the variability of the preparations of the ensemble, greater variability means the ensemble is spread over more cases. Therefore, the count of distinguishable states must be a monotonic function of the entropy. It has been shown[cite Hall] under very general conditions that the exponential of the entropy is the only function to have the appropriate features. One required feature, that we prove in general, is that the exponential of the entropy is subadditive (i.e. the count of states associated to a mixture cannot exceed the combined count of states associated to both mixtures) and it is additive for orthogonal ensembles (i.e. no overlap in possible instances) in a specific linear combination.

@@

The first property readily comes from the range of the exponential. For the second, note that $\hull$, exponential, and $\sup$ are all monotone. For sub-additivity, the exponential of the entropy itself is sub-additive, and it is additive over orthogonal ensembles. For the last, since the entropy itself is continuous, this property carries over to the set function.

Again, we note that the non-additive measure we recover is very well-behaved and connects directly with important quantities. When Choquet theory applies, state capacity can be extended to the extreme points. In the classical case, disjoint sets of extreme points will be orthogonal, meaning that the state capacity there restricted is an additive measure. In the discrete case, every point is associated with zero entropy, and since $2^0=1$, the state capacity reduces to the counting measure. In the continuous case, given a subset of phase space $A$, the maximum entropy is reached by the uniform distribution $\rho_A$, which means we recover $\log_2 \capacity(A) = S(\rho_A)$, the fundamental postulate of statistical mechanics. Similarly, in quantum mechanics, given an orthogonal subspace $A$, the maximum entropy is recovered by the maximally mixed state, which means $\capacity(A) = 2^{\dim(A)}$.

The state capacity, then, recovers the number of distinguishable states. In classical mechanics, all states are perfectly distinguishable. In quantum mechanics, only orthogonal states are distinguishable, and therefore the state capacity recovers the number of orthogonal states.
